---
title: "Tasks and Templates"
author: "Michal Burda"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tasks and Templates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r tt-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette covers advanced features of `rmake` that help manage complex build processes: **tasks** for organizing and selectively executing groups of rules, and **templates** for efficiently generating multiple similar rules with parameterization.

For basic usage and an introduction to rmake, see the **Getting Started with rmake** vignette. For information on specific rule types, see the **Build Rules** vignette. For project management basics, see the **rmake Project Management** vignette.

## Tasks

Tasks allow grouping rules that can be executed together. Each rule is a member of the `"all"` task by default. Rules can belong to multiple tasks.

### Executing Tasks

Execute a task:
```{r task_ex, eval=FALSE}
make('all')
make('preview')
```

### Assigning Rules to Tasks

Assign rules to tasks:
```{r task_assign, eval=FALSE}
library(rmake)
job <- c(
  "data.csv" %>>% rRule("preprocess.R") %>>% "data.rds",
  "data.rds" %>>% markdownRule("preview.Rmd", task = "preview") %>>% 
    "preview.pdf",
  "data.rds" %>>% markdownRule("final.Rmd", task = "final") %>>% 
    "final.pdf"
)
makefile(job, "Makefile")
```

Running `make("preview")` creates `data.rds` and `preview.pdf` but not `final.pdf`.

### Use Cases for Tasks

Tasks are useful for:

- **Incremental development**: Create a "preview" task for quick iterations and a "final" task for complete analysis
- **Different outputs**: Separate tasks for different report formats or audiences
- **Testing**: Create a "test" task that runs on sample data before processing the full dataset

## Parameterized Execution

Pass parameters to scripts via the `params` argument:

```{r params_ex, eval=FALSE}
library(rmake)
job <- c(
  "data.csv" %>>% rRule("fit.R", params = list(alpha = 0.1)) %>>% "out-0.1.rds",
  "data.csv" %>>% rRule("fit.R", params = list(alpha = 0.2)) %>>% "out-0.2.rds",
  "data.csv" %>>% rRule("fit.R", params = list(alpha = 0.3)) %>>% "out-0.3.rds"
)
makefile(job, "Makefile")
```

### Accessing Parameters in Scripts

Parameters are available in scripts as the `params` global variable:
```{r params_script, eval=FALSE}
# fit.R
str(params)
# List of 5
#  $ .target : chr "out-0.1.rds"
#  $ .script : chr "fit.R"
#  $ .depends: chr "data.csv"
#  $ .task   : chr "all"
#  $ alpha   : num 0.1
```

### Using getParam()

Use `getParam()` to access parameters safely:
```{r getparam_ex, eval=FALSE}
# fit.R
library(rmake)

dataName <- getParam(".depends")
resultName <- getParam(".target")
alpha <- getParam("alpha")

# Use parameters...
cat("Processing with alpha =", alpha, "\n")
```

`getParam()` with default values:
```{r getparam_default, eval=FALSE}
dataName <- getParam(".depends", "data.csv")
resultName <- getParam(".target", "result.rds")
alpha <- getParam("alpha", 0.2)
```

### Built-in Parameters

Every rule automatically has the following parameters available:

- `.target` - The target file(s) being created
- `.script` - The script being executed
- `.depends` - The dependency file(s)
- `.task` - The task name(s)

## Rule Templates

Rule templates avoid repetitive rule definitions using `expandTemplate()`. This is particularly useful when you need to run the same analysis with different parameters or on different datasets.

### Simple Template

```{r template_simple, eval=FALSE}
tmpl <- "data-$[NUM].csv" %>>% 
  rRule("process.R") %>>% 
  "result-$[NUM].csv"
variants <- data.frame(NUM = 1:99)
job <- expandTemplate(tmpl, variants)
```

This creates 99 rules, one for each value of `NUM`.

### Template with Multiple Variables

```{r template_multi}
variants <- expand.grid(DATA = c("dataSimple", "dataComplex"),
                        TYPE = c("lm", "rf", "nnet"))
print(variants)

tmpl <- "$[DATA].csv" %>>% 
  rRule("fit-$[TYPE].R") %>>%
  "result-$[DATA]_$[TYPE].csv"
job <- expandTemplate(tmpl, variants)
print(job)
```

### Combining Templates with Parameters

Duplicate rules are automatically removed:
```{r template_dup}
tmpl <- "data.csv" %>>%
  rRule("pre.R") %>>% "pre.rds" %>>%
  rRule("comp.R", params = list(alpha = "$[NUM]")) %>>% 
  "result-$[NUM].csv"
variants <- data.frame(NUM = 1:5)
job <- expandTemplate(tmpl, variants)
print(job)
```

### Template Variables in Parameters

Template variables (like `$[NUM]`) can be used in:
- File names (targets and dependencies)
- Script names
- Parameters passed to rules

### Common Pitfall: Multiple Rules for Same Target

Warning: Different rules producing the same target will cause an error:
```{r template_error, error=TRUE}
tmpl <- "data-$[TYPE].csv" %>>% 
  markdownRule("report.Rmd") %>>% "report.pdf"
variants <- data.frame(TYPE = c("a", "b", "c"))
job <- expandTemplate(tmpl, variants)
print(job)

# This would error:
# makefile(job)
# Error: Multiple rules detected for the same target
```

## Combining Tasks and Templates

You can combine tasks and templates for powerful workflow management:

```{r combined_ex, eval=FALSE}
# Create template for different models
tmpl <- "data.csv" %>>% 
  rRule("fit-$[MODEL].R", params = list(model = "$[MODEL]")) %>>%
  "model-$[MODEL].rds"

# Expand for different models
variants <- data.frame(MODEL = c("linear", "rf", "xgboost"))
training_job <- expandTemplate(tmpl, variants)

# Add evaluation tasks
eval_job <- c(
  c("model-linear.rds", "model-rf.rds", "model-xgboost.rds") %>>%
    rRule("evaluate.R") %>>% "evaluation.rds",
  "evaluation.rds" %>>% 
    markdownRule("report.Rmd", task = "report") %>>% "report.pdf"
)

# Combine all rules
job <- c(training_job, eval_job)
makefile(job, "Makefile")
```

## Best Practices

### For Tasks

1. **Use meaningful names**: Choose task names that clearly describe their purpose
2. **Keep "all" minimal**: Only include essential outputs in the default "all" task
3. **Create incremental tasks**: Build up complexity with tasks like "quick", "full", "validate"

### For Templates

1. **Use data frames for variants**: This makes it easy to generate all combinations with `expand.grid()`
2. **Test with small variants first**: Before generating hundreds of rules, test with 2-3 variants
3. **Use consistent naming**: Develop a naming convention for template variables
4. **Document your templates**: Add comments explaining what each template generates

## Summary

This vignette covered:

- **Tasks**: Organizing rules into groups that can be executed selectively
- **Parameterized Execution**: Passing parameters to scripts for flexible processing
- **Rule Templates**: Efficiently generating multiple similar rules
- **Combining Features**: Using tasks and templates together for complex workflows

For more information on basic usage and rule types, see the other vignettes:
- "Getting Started with rmake"
- "rmake Project Management"
- "Build Rules"
